---
title: "Investigating Factors That Influence High IMDB Film Ratings"
number-sections: true
format: 
  html:
    embed-resources: true
    code-tools: true
editor_options:
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: libraries

library(tidyverse)
library(gt)
library(gridExtra)
library(moderndive)
dataset09 <- read_csv("https://raw.githubusercontent.com/Janet180112/DAS-Group-09/main/dataset09.csv")
```
# Data Preparation {#sec-data-prep}
Before starting the analysis, the dataset was checked for missing values. 
The results show that the variable length has 127 missing values. 
These rows were removed before the analysis.

```{r}

# Check missing values
colSums(is.na(dataset09))

# Remove rows with missing values
dataset09 <- na.omit(dataset09)
# Summary statistics
str(dataset09)

```

# Introduction {#sec-Intro}
Movies are a common form of entertainment, and many people look at online ratings before choosing what to watch. IMDB is a well-known website where viewers rate films on a scale from 0 to 10. However, not all films receive high ratings, and different features of a film may influence how audiences rate them. In this project, we use an IMDB dataset to investigate which film characteristics affect whether a film receives a rating higher than 7.

# Exploratory Data Analysis {#sec-EDA}

# Formal Data Analysis {#sec-FDA}

# Conclusions {#sec-Conc}


